Exploring the composition of lithic assemblages in Mesolithic south-eastern Norway

### 2021-05-05 analysis/paper/paper.Rmd
Set up the function for modelling radiocarbon dates and summing the posterior distributions. Struggling with ordering the dates properly for the plot

Even though each individual assemblage can have been impacted by an infinitude of effects that might skew an archaeological interpretation, this does not preclude the applicability of inductive analyses aimed at revealing overarching structure in the data, without imposing overly complex analytical frameworks that attempt to account for these particularities [@bevan2015]. Structure that can be revealed from considering all of the assemblages in aggregate can constitute a step in an iterative analytical chain that ultimately aims to tease apart such aspects from the multitude of factors that have shaped the composition of the assemblages, and should be of value to subsequent in-depth studies of any individual site. The most immediate danger of the approach outlined here is rather to be overly naive in the causal significance and cultural importance that is ascribed to any identified pattern. As such, the main aim of this analysis is to compare the results with findings reported in previous literature concerned with the Mesolithic in southern Norway, that are typically based on more informally driven methods, and have the generation of new hypotheses as a possible outcome. To this end, the analysis follows two analytical avenues. The first involves an analysis of the assemblages that is based on the classification of the artefacts done for the original excavation reports. The second involves an analysis of the assemblages in light of the whole assemblage behaviour index, which has not seen application in this context before.

@@

```{r wrangling}
sites <- assemblages %>% 
      # Find mean reported date for each site (to be used in the plots below)
  rowwise() %>% 
  mutate(mean_date = mean(c(reported_start_bp, reported_end_bp))) %>% 
  mutate(
    # Find total artefact count and flint count for each site
    artefact_count = sum(c_across(core:flake_retouch_other)),
    flint_count = sum(c_across(core:flake_retouch)),
    nonflint_count = sum(c_across(axe_other:flake_retouch_other))) %>%
  # Exclude sites with less than 100 artefacts
  filter(artefact_count > 100)

h3 <- radiocarbon %>% 
  filter(site == "Vallermyrene 1")

@@

oxcal_code <- paste0('Plot()
   {
    Sequence()
    {
     Boundary("Start");
     Phase()
     {
    ',
      h3dat,
     '
     Sum("Vallermyrene 1");
     };
     Boundary("End");
    };
  };')
my_result_file <- executeOxcalScript(oxcal_code)
my_result_text <- readOxcalOutput(my_result_file)
my_result_data <- parseOxcalOutput(my_result_text, only.R_Date = FALSE)
plot(my_result_data$` Vallermyrene 1 `$posterior_probabilities$dates, my_result_data$` Vallermyrene 1 `$posterior_probabilities$probabilities, type = "l")

@@

yrs <- seq(11500, 5000, -500)
temp_dist <- ggplot(sites, aes(x = yrs, y = reorder(site, mean_date))) +
  geom_segment(aes(x = reported_start_bp, xend = reported_end_bp,
                   yend = reorder(site, mean_date),
                   colour = excavation_project), size = 1.7) +
  xlab("Cal. BP") +
  scale_x_reverse() +
  scale_colour_manual(labels = c("E18 Bommestad-Sky", "E18 Brunlanes",
                               "E18 Rugtvedt-Dørdal", "Vestfoldbaneprosjektet"),
                     values = c("springgreen4", "royalblue2",
                                "gold1", "firebrick3")) +
  labs(title = "B", colour = "Excavation project") +
  theme_classic() +
  theme(axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        legend.position = "bottom") +
  guides(colour =  guide_legend(title.position = "top", title.hjust = 0.5))

@@

```{r co_data}
# Perform correspondence analysis on the data. 'mean_age' is meant for colouring
# so is excluded here
calithics <- CA(select(lithics, -mean_date, -artefact_count, -flint_count, 
                       -nonflint_count, -reported_start_bp, -reported_end_bp,
                       -askeladden_id),
                graph = FALSE)

@@

lithics2 <- sites %>% 
  select(-n, -beach_flint_flint_nodule, -hammerstone, -excavation_project,
         -reported_start_bp, -reported_end_bp, -askeladden_id) %>% 
  # The reason for not using .keep below is for the option to 
  #retain artefact_count
  mutate(chip_freq =  sum(chip, chip_other) / artefact_count,
         core_freq = sum(core, core_other) / artefact_count,
         blanks_freq = sum(blade, microblade, fragment, flake, flake_other, 
                     fragment_other, blade_other,  microblade_other) / artefact_count,
         secondary_freq =  sum(axe_debitage, axe, chisel, 
                microlith, tanged_point, transverse_point, single_edged_point,
                hognipen_point, projectile_undefined, microburin, scraper, 
                knife, burin, drill, blade_retouch, microblade_retouch, 
                chip_retouch, fragment_retouch, flake_retouch, axe_other,
                axe_debitage_other, chisel_other, club_hatchet_other, 
                grinding_slab_stone, projectile_point_other, knife_other,
                scraper_other, drill_other, microblade_retouch_other, 
                blade_retouch_other, fragment_retouch_other,
                flake_retouch_other) / artefact_count,
         nonflint_freq = nonflint_count / artefact_count,
         lithic_dens = artefact_count / excavated_m3) %>%
  select(-blanks_lithics, -secondary_lithics, -excavated_m3, -artefact_count,
         -flint_count, -nonflint_count, -core, -core_fragment, -chip, 
         -chip_other, -core_other, -core_fragment_other, -dating_method) %>% 
  column_to_rownames("site")

# Repeat but for flint only
flint2 <- sites %>% 
  select(-n, -beach_flint_flint_nodule, -hammerstone, -excavation_project, 
         -reported_start_bp, -reported_end_bp, -askeladden_id) %>% 
  select(!ends_with("other")) %>% 
  mutate(chip_freq =  sum(chip) / artefact_count,
         core_freq = sum(core) / artefact_count,
         blanks_freq = sum(blade, microblade, fragment, flake) / artefact_count,
         secondary_freq =  sum(axe_debitage, axe, chisel, 
                microlith, tanged_point, transverse_point, single_edged_point,
                hognipen_point, projectile_undefined, microburin, scraper, 
                knife, burin, drill, blade_retouch, microblade_retouch, 
                chip_retouch, fragment_retouch, flake_retouch) / artefact_count,
         flint_dens = flint_count / excavated_m3) %>%
  select(-blanks_flint, -secondary_flint, -excavated_m3, -artefact_count,
         -flint_count, -nonflint_count, -grinding_slab_stone, -core, 
         -core_fragment, -chip, dating_method) %>% 
  column_to_rownames("site")

@@

```{r cor-plot, fig.cap = "Correlation plot showing transformation of skewed variables for the PCA, and the linear correlation between these", fig.height = 9.5, fig.width = 10}
# Transform the skewed variables - sqrt() is used where there are zero values,
# and seems to provide an adequate result.
lithicstr <- lithics2 %>%
  # Transform skewed variables and reorder
  mutate("Mean date cal. BP*-1" = mean_date * -1,
          "log10(lithic_dens)" = log10(lithic_dens),
         "log10(secondary_freq)" = log10(secondary_freq),
         "core_freq^(1/3)" = core_freq^(1/3),
         "nonflint_freq^(1/3)" = nonflint_freq^(1/3),  .keep = "unused") %>% 
  relocate("Mean date cal. BP*-1",
    "log10(lithic_dens)", "blanks_freq", "log10(secondary_freq)",
    "chip_freq", "core_freq^(1/3)" 
  )

@@

```{r pca-analysis, fig.cap = "PCA using variables related. Note that details on the transformation of the variables has been left out of the plot for clarity.", fig.height = 7, fig.width = 10}
# Rename the transformed variables again to avoid clutter in the plot. 
# Specify in figure caption that the transformed variables are used 
lithicspca <- lithicstr %>% 
  mutate(mean_date =  `Mean date cal. BP*-1` * -1, .keep = "unused") %>% 
  rename(
   lithic_dens = "log10(lithic_dens)",
   secondary_freq = "log10(secondary_freq)",
   core_freq = "core_freq^(1/3)",
   nonflint_freq = "nonflint_freq^(1/3)"
  )

@@

w3 <- ggplot(indexdata, aes(x = mean_date, y = index)) + 
  geom_point() +
  labs(title = "C",
       x = "Mean date cal. BP",
       y = "Curation index") +
  scale_x_reverse() + 
  geom_smooth(method = "loess", col = "red") +
  theme_classic() 
w3
```
# Discussion

>>>>

Even though each individual assemblage can have been impacted by an infinitude of effects that might skew an archaeological interpretation, this does not preclude the applicability of inductive analyses aimed at revealing overarching structure in the data, without imposing overly complex analytical frameworks that attempt to account for these particularities [@bevan2015]. Structure that can be revealed from considering all of the assemblages in aggregate can constitute a step in an iterative analytical chain that ultimately aims to tease apart such aspects from the multitude of factors that have shaped the composition of the assemblages, and should be of value to subsequent in-depth studies of any individual site. The most immediate danger of the approach outlined here is rather to be overly naive in the causal significance and cultural importance that is ascribed to any identified pattern. As such, the main aim of this analysis is to compare the results with findings reported in previous literature concerned with the Mesolithic in southern Norway and have the generation of new hypotheses as a possible outcome. To this end, the analysis follows two analytical avenues. The first involves an analysis of the assemblages that is based on the classification of the artefacts done for the original excavation reports. The second involves an analysis of the assemblages in light of the whole assemblage behaviour index, which has not seen application in this context before.

@@

```{r wrangling}
sites <- assemblages %>% 
      # Find mean reported date for each site (to be used in the plots below)
  rowwise() %>% 
  mutate(mean_date = mean(c(reported_start_bc, reported_end_bc))) %>% 
  mutate(
    # Find total artefact count and flint count for each site
    artefact_count = sum(c_across(core:flake_retouch_other)),
    flint_count = sum(c_across(core:flake_retouch)),
    nonflint_count = sum(c_across(axe_other:flake_retouch_other))) %>%
  # Exclude sites with less than 100 artefacts
  filter(artefact_count > 100)
```

```{r c14-modelling}
# define function to model radiocarbon dates. Only assumption is that 
# dates from a single site belong to a single group
model_rcarb <- function(data){

  # Format individual dates for R_Date in Oxcal
  dates <- R_Date(data[["lab_code"]], data[["c14_bp"]], data[["error"]])

  # Manually compile rest of OxCal code
  oxcal_code <- paste0('
  Plot()
   {
    Sequence()
    {
     Boundary("Start");
     Phase()
     {
    ',
     dates,
     '
     Sum("Sum");
     };
     Boundary("End");
    };
  };')

  # Run OxCal code - wrapped this in some functions to  not print to 
  # rmarkdown
  invisible(capture.output(oxcal_exec <- executeOxcalScript(oxcal_code)))
  oxcal_read <- readOxcalOutput(oxcal_exec)
  # Note that parseOxcalOutput() defaults to only returning individual dates
  oxcal_data <- parseOxcalOutput(oxcal_read, only.R_Date = FALSE)

  # Assemble data frame to be returned
  posterior <- data.frame(
    probabilities = oxcal_data$Sum$posterior_probabilities$probabilities,
    date_grid = oxcal_data$Sum$posterior_probabilities$dates,
    site = data[["site"]][1],
    excavation_project = data[["excavation_project"]][1])

  # Return data
  return(posterior)
}

# Apply modelling function to the radiocarbon dates, grouped by site name
posteriors <- radiocarbon %>% 
  group_by(site) %>%
  group_map(~ model_rcarb(.), .keep = TRUE)

# The data is returned as a list of lists that needs to be unpacked
rcarb_data <- do.call(rbind, posteriors) 
  group_by(site) %>%
  mutate(meandate = mean(date_grid * -1)) %>% 
  ungroup() %>% 
  arrange(meandate)

h3 <- radiocarbon %>% 
  filter(site == "Vallermyrene 1" | site == "Stokke/Polland 5")

v1 <- model_rcarb(h3)

@@

my_result_file <- executeOxcalScript(oxcal_code)
my_result_text <- readOxcalOutput(my_result_file)
my_result_data <- parseOxcalOutput(my_result_text, only.R_Date = FALSE)
plot(my_result_data$`Vallermyrene 1`$posterior_probabilities$dates, my_result_data$`Vallermyrene 1`$posterior_probabilities$probabilities, type = "l")

posterior <- data.frame(probabilities = my_result_data$`Vallermyrene 1`$posterior_probabilities$probabilities,
                        date_grid = my_result_data$`Vallermyrene 1`$posterior_probabilities$dates,
                        site = "Vallermyrene 1",
                        excavation_project = "Vestfoldbaneprosjektet")

ggplot() +
  geom_line(aes(x = yrs, y = my_result_data$` Vallermyrene 1 `$posterior_probabilities$probabilities))

yrs <- seq(11500, 5000, -500)
plt1 <- ggplot(sites, aes(x = yrs)) +
  geom_segment(aes(x = reported_start_bc, xend = reported_end_bc,
                   yend = reorder(site, mean_date), y = reorder(site, mean_date),
                   colour = excavation_project), size = 1.7) +
  xlab("Cal. BC") +
  scale_colour_manual(labels = c("E18 Bommestad-Sky", "E18 Brunlanes",
                               "E18 Rugtvedt-Dørdal", "Vestfoldbaneprosjektet"),
                     values = c("springgreen4", "royalblue2",
                                "gold1", "firebrick3")) +
  labs(title = "B", colour = "Excavation project") +
  theme_classic() +
  theme(axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        legend.position = "bottom") +
  guides(colour =  guide_legend(title.position = "top", title.hjust = 0.5)) +
    scale_x_reverse()

ggplot(sites, aes(x = yrs)) +
      ggridges::geom_ridgeline(data = rcarb_data, aes(x = date_grid * -1,
                                                      y = fct_reorder(site, date_grid * -1, .fun = mean),
                                      height = probabilities * 100,
                                      fill = excavation_project)) +
    geom_segment(aes(x = reported_start_bc, xend = reported_end_bc,
                   yend = reorder(site, mean_date), y = reorder(site, mean_date),
                   colour = excavation_project), size = 1.7) +
  xlab("Cal. BC") +
  scale_colour_manual(labels = c("E18 Bommestad-Sky", "E18 Brunlanes",
                               "E18 Rugtvedt-Dørdal", "Vestfoldbaneprosjektet"),
                     values = c("springgreen4", "royalblue2",
                                "gold1", "firebrick3")) +
  labs(title = "B", colour = "Excavation project") +
  theme_classic() +
  theme(axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        legend.position = "bottom") +
  guides(colour =  guide_legend(title.position = "top", title.hjust = 0.5)) +
    scale_x_reverse()

    ggridges::geom_ridgeline(data = rcarb_data, aes(x = date_grid * -1,
                                      y = fct_reorder(as.factor(site), date_grid * -1, .fun = mean),
                                      height = probabilities * 100,
                                      group = fct_reorder(as.factor(site), date_grid * -1, .fun = mean),
                                      fill = excavation_project)) +
  scale_fill_manual(labels = c("E18 Rugtvedt-Dørdal", 
                                "Vestfoldbaneprosjektet", "E18 Bommestad-Sky"),
                     values = c("springgreen4",
                                "gold1", "firebrick3"),
                    guide = FALSE) +

@@

yrs <- seq(11500, 5000, -500)
temp_dist <- ggplot(sites, aes(x = yrs, y = reorder(site, mean_date))) +
  geom_segment(aes(x = reported_start_bc, xend = reported_end_bc,
                   yend = reorder(site, mean_date),
                   colour = excavation_project), size = 1.7) +
  xlab("Cal. BC") +
  scale_x_reverse() +
  scale_colour_manual(labels = c("E18 Bommestad-Sky", "E18 Brunlanes",
                               "E18 Rugtvedt-Dørdal", "Vestfoldbaneprosjektet"),
                     values = c("springgreen4", "royalblue2",
                                "gold1", "firebrick3")) +
  labs(title = "B", colour = "Excavation project") +
  theme_classic() +
  theme(axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        legend.position = "bottom") +
  guides(colour =  guide_legend(title.position = "top", title.hjust = 0.5))

@@

```{r co_data}
# Perform correspondence analysis on the data. 'mean_age' is meant for colouring
# so is excluded here
calithics <- CA(select(lithics, -mean_date, -artefact_count, -flint_count, 
                       -nonflint_count, -reported_start_bc, -reported_end_bc,
                       -askeladden_id),
                graph = FALSE)

@@

lithics2 <- sites %>% 
  select(-n, -beach_flint_flint_nodule, -hammerstone, -excavation_project,
         -reported_start_bc, -reported_end_bc, -askeladden_id) %>% 
  # The reason for not using .keep below is for the option to 
  #retain artefact_count
  mutate(chip_freq =  sum(chip, chip_other) / artefact_count,
         core_freq = sum(core, core_other) / artefact_count,
         blanks_freq = sum(blade, microblade, fragment, flake, flake_other, 
                     fragment_other, blade_other,  microblade_other) / artefact_count,
         secondary_freq =  sum(axe_debitage, axe, chisel, 
                microlith, tanged_point, transverse_point, single_edged_point,
                hognipen_point, projectile_undefined, microburin, scraper, 
                knife, burin, drill, blade_retouch, microblade_retouch, 
                chip_retouch, fragment_retouch, flake_retouch, axe_other,
                axe_debitage_other, chisel_other, club_hatchet_other, 
                grinding_slab_stone, projectile_point_other, knife_other,
                scraper_other, drill_other, microblade_retouch_other, 
                blade_retouch_other, fragment_retouch_other,
                flake_retouch_other) / artefact_count,
         nonflint_freq = nonflint_count / artefact_count,
         lithic_dens = artefact_count / excavated_m3) %>%
  select(-blanks_lithics, -secondary_lithics, -excavated_m3, -artefact_count,
         -flint_count, -nonflint_count, -core, -core_fragment, -chip, 
         -chip_other, -core_other, -core_fragment_other, -dating_method) %>% 
  column_to_rownames("site")

# Repeat but for flint only
flint2 <- sites %>% 
  select(-n, -beach_flint_flint_nodule, -hammerstone, -excavation_project, 
         -reported_start_bc, -reported_end_bc, -askeladden_id) %>% 
  select(!ends_with("other")) %>% 
  mutate(chip_freq =  sum(chip) / artefact_count,
         core_freq = sum(core) / artefact_count,
         blanks_freq = sum(blade, microblade, fragment, flake) / artefact_count,
         secondary_freq =  sum(axe_debitage, axe, chisel, 
                microlith, tanged_point, transverse_point, single_edged_point,
                hognipen_point, projectile_undefined, microburin, scraper, 
                knife, burin, drill, blade_retouch, microblade_retouch, 
                chip_retouch, fragment_retouch, flake_retouch) / artefact_count,
         flint_dens = flint_count / excavated_m3) %>%
  select(-blanks_flint, -secondary_flint, -excavated_m3, -artefact_count,
         -flint_count, -nonflint_count, -grinding_slab_stone, -core, 
         -core_fragment, -chip, dating_method) %>% 
  column_to_rownames("site")

@@

```{r cor-plot, fig.cap = "Correlation plot showing transformation of skewed variables for the PCA, and the linear correlation between these", fig.height = 9.5, fig.width = 10}
# Transform the skewed variables - sqrt() is used where there are zero values,
# and seems to provide an adequate result.
lithicstr <- lithics2 %>%
  # Transform skewed variables and reorder
  mutate("Mean date cal. BC*-1" = mean_date * -1,
          "log10(lithic_dens)" = log10(lithic_dens),
         "log10(secondary_freq)" = log10(secondary_freq),
         "core_freq^(1/3)" = core_freq^(1/3),
         "nonflint_freq^(1/3)" = nonflint_freq^(1/3),  .keep = "unused") %>% 
  relocate("Mean date cal. BC*-1",
    "log10(lithic_dens)", "blanks_freq", "log10(secondary_freq)",
    "chip_freq", "core_freq^(1/3)" 
  )

@@

```{r pca-analysis, fig.cap = "PCA using variables related. Note that details on the transformation of the variables has been left out of the plot for clarity.", fig.height = 7, fig.width = 10}
# Rename the transformed variables again to avoid clutter in the plot. 
# Specify in figure caption that the transformed variables are used 
lithicspca <- lithicstr %>% 
  mutate(mean_date =  `Mean date cal. BC*-1` * -1, .keep = "unused") %>% 
  rename(
   lithic_dens = "log10(lithic_dens)",
   secondary_freq = "log10(secondary_freq)",
   core_freq = "core_freq^(1/3)",
   nonflint_freq = "nonflint_freq^(1/3)"
  )

@@

w3 <- ggplot(indexdata, aes(x = mean_date, y = index)) + 
  geom_point() +
  labs(title = "C",
       x = "Mean date cal. BC",
       y = "Curation index") +
  scale_x_reverse() + 
  geom_smooth(method = "loess", col = "red") +
  theme_classic() 
w3
```
# Discussion
